---
title: "Darwin Core mapping"
subtitle: "For dataset: Checklist of the rust fungi of Belgium"
author: 
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
date:"`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

This document describes how we map the checklist data to Darwin Core.

# Setup

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r, eval = F}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # For data transformations

# None core tidyverse packages:
library(magrittr)  # For %<>% pipes

# Other packages
library(janitor)   # For cleaning input data
library(digest)    # Create hash function digests
library(readr)     # To read a delimited file into a tibble
library(rgbif)     # Interface to the GBIF API
library(stringi)   # For fast and convenient string manipulation
library(stringr)   # For fast and convenient string manipulation
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/checklist-of-rusts.tsv"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
```

# Read data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_delim(raw_data_file, delim = "\t") 

# Clean the data somewhat: remove empty rows if present
raw_data %<>%
  remove_empty_rows()     # Remove empty rows
```

## Extract host plant information

The checklist includes information about rust fungi (`scientificName`) and their associated host plants (`hostPlant`). Although the publication of belgian rust fungi is our central focus, it is also informative to include the host taxa in the taxon core. In the [Resource Relation Extension]() (see paragraph xxx), we elaborate further on the relation between the specific host and plant species.

Before we start the mapping process, we need to extract the host plant information from `hostPlant` in `raw data` and we add them to the checklist as extra species under `scientificName`.

Extract host plant and rust fungi information from `raw_data` (we also include the rust fungi as an intermediate step, as we need the combined information later for the mapping of the Resource Relationship extension):

```{r}
host_plant <- raw_data %>% select(scientificName, hostPlant)
```

```{r, echo = FALSE}
head(host_plant)
```

Split `hostPlant` on scientific name, using "," as a separator:

```{r}
host_plant %<>% separate(
  hostPlant, 
  sep = ", ", 
  into = c(paste("species", c(1:20), sep = "_")),
  fill = "right")
```

Gather all host species in one row (omitting "NA"):

```{r}
host_plant %<>% gather("interaction", "scientificName_host", starts_with("species"), na.rm = TRUE)
```

Remove leading whitespaces:

```{r}
host_plant %<>% mutate(scientificName_host = str_trim(scientificName_host))
```

This dataframe gives a nice structured overview of all interactions between the host plant and the parasite. We save this dataframe as we will need this to construct the Resource Relationship extension (see xxx):

```{r}
interactions <- host_plant
```

Select unique scientificNames:

```{r}
host_plant %<>% distinct(scientificName_host) %<>% 
  arrange(scientificName_host)
```

Add a column `resourceRelation` to `raw_data` and `host_plant` to indicate whether the species in `scientificName`is a parasite or a host:

```{r}
raw_data %<>% mutate(resourceRelation = "parasite")
host_plant %<>% mutate(resourceRelation = "host")
```

Bind `raw_data` and `host_plant` by rows:

```{r}
raw_data %<>% bind_rows(host_plant)
```

## Generate taxonID

To uniquely identify a taxon in the Taxon Core and reference taxa in the Extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable: 

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("uredinales-belgium-checklist", "taxon", vdigest(scientificName, algo="md5"), sep=":"))
```

Create a data frame `raw_data` from the source data:

```{r}
# Add prefix `raw_` to all column names in `raw_data`:
colnames(raw_data) <- paste0("raw_", colnames(raw_data))

# Save those column names as a vector (makes it easier to remove them all later):
raw_colnames <- colnames(raw_data)
```

Preview data:

```{r}
head(raw_data)
```

## Add bibliographicCitation

This checklist is a compilation of three published volumes of the `Catalogue des Uredinales de Belgique (The rust fungi of Belgium)`. For each rust fungus, a volume number (`raw_part`) and a page number (`raw_page`) is provided. This provides a concise description of the species, most of the information is about taxonomy, observations and the hosts. We need to integrate these citations in both the Taxon Core and the Extensions. So, we add `bibliographicCitation` to `raw_data`. 

The `bibliographicCitation` will contain the full reference with page number. First, we complement the volume number with the full reference (given in the metadata). For this, we create the data frame `full_reference`:

```{r, echo = F}
(full_reference <- as.data.frame(matrix(
  c(1, "Vanderweyen, A., & Fraiture, A. (2009). Catalogue des Uredinales de Belgique, 1re partie, Chaconiaceae, Coleosporiaceae, Cronartiaceae, Melampsoraceae, Phragmidiaceae, Pucciniastraceae, Raveneliaceae et Uropyxidaceae. Lejeunia, Revue de Botanique.",
    2, "Vanderweyen, A., & Fraiture, A. (2009). Catalogue des Uredinales de Belgique, 2ème partie, Pucciniaceae (sauf Puccinia)(suite 2). Lejeunia, Revue de Botanique.",
    3, "Vanderweyen, A., & Fraiture, A. (2012). Catalogue des Uredinales de Belgique, 3ème partie, Pucciniaceae (genre Puccinia). Lejeunia, Revue de Botanique."), 
  nrow = 3, 
  ncol = 2, 
  byrow = T,
  dimnames = list (c(1:3), c("raw_part", "full_reference")))))

```

To join both data frames, `raw_part` in `full_reference` should be an integer:

```{r}
full_reference %<>% mutate(raw_part = as.integer(raw_part))
```

Join both data frames:

```{r}
raw_data <- left_join(raw_data, full_reference, by = "raw_part")
```

For each rust fungus taxon, the `bibliographicCitation` will contain the full reference with page number. For all host species, we refer to "Vanderweyen, A., & Fraiture, A. (2009, 2012) Catalogue des Uredinales de Belgique" (excluding volumne and page number). This is because the complete reference (with volumne and page number) applies to the rust fungus, not the host plant. One host plant taxon can be infected by multiple rust fungi taxa and would thus contain multipele complete references.

```{r}
raw_data %<>% mutate(raw_bibliographicCitation = case_when(
  raw_resourceRelation == "parasite" ~  paste(full_reference, "Page:", raw_page),
  raw_resourceRelation == "host"     ~ "Vanderweyen, A., & Fraiture, A. (2009, 2012) Catalogue des Uredinales de Belgique"))
```

Remove `full_reference`:

```{r}
raw_data %<>% select(-full_reference)
```

# Create Taxon core

## Pre-processing

```{r}
taxon <- raw_data
```

### generic names

`raw_scientificName` includes information about the genus, specific epithet, infraspecific epithet, (bracket)authorship and taxonRank. The full scientific name will be mapped under `scientificName` in the Taxon Core. For enhanced readability, we parse `raw_scientificName` into its different components using the parsenames() function from [rgbif](https://cran.r-project.org/web/packages/rgbif/rgbif.pdf).

```{r}
parsed_names <- parsenames(taxon $ raw_scientificName)
```

Overview of the different compenents:

```{r}
head(parsed_names)
```

Select the columns needed for further mapping in the Taxon Core, i.e. `genusorabove`, `specificepithet`, `authorship`, `bracketauthorship`, `rankmarker` and `infraspecificepithet` (these will be mapped to, respectively, `genus`, `specificEpithet`, `scientificNameAuthorship`, `scientificNameAuthorship`, `taxonRank` and `infraspecificEpithet`). We also integrated `parsed`. If `parsed` = `FALSE`, than the scientificName could not by parsed by GBIF. We also select `scientificname` as we need this to merge `parsed_names` with `scientificName` in the Taxon Core. 

```{r}
parsed_names %<>% select(scientificname, genusorabove, specificepithet, authorship, bracketauthorship, rankmarker, infraspecificepithet, parsed)
```

Some columns contain NA's. We replace this by blanks:

```{r}
parsed_names %<>% replace_na(list(
  specificepithet      = "",   
  infraspecificepithet = "",
  authorship           = "",
  bracketauthorship    = "",
  rankmarker           = ""))
```

Similar as for the raw data file, we add the prefix `pn_` to the column names of `parsed_names`. This is because we merge `parsed_names` to the Taxon Core and remove these columns after mapping to Darwin Core terms. 

```{r}
colnames(parsed_names) <- paste0("pn_", colnames(parsed_names))

# Save those column names as a vector (makes it easier to remove them all later):
pn_colnames <- colnames(parsed_names)
```

Amount of species in the Taxon Core before merge:

```{r, echo = F}
(before_merge <- nrow(taxon))
```

Merge `parsed_names` with the Taxon Core:

```{r}
taxon <- inner_join(taxon, parsed_names, by = c("raw_scientificName" = "pn_scientificname"))
```

Is the amount of species in the Taxon Core the same as before the merge?

```{r, echo = F}
before_merge == nrow(taxon)
```

## Term mapping
 
Map the source data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml):
 
### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "Botanical Garden Meise")

```

### bibliographicCitation

```{r}
taxon %<>% mutate(bibliographicCitation = raw_bibliographicCitation) 
```


### institutionID

```{r}
taxon %<>% mutate(institutionID = "http://biocol.org/urn:lsid:biocol.org:col:15605")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Checklist of the Rust Fungi of Belgium")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_scientificName)
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = case_when(
  raw_resourceRelation == "parasite" ~ "Fungi",
  raw_resourceRelation == "host"     ~ "Plantae"))
```

### phylum

```{r}
taxon %<>% mutate(phylum = case_when(
  raw_resourceRelation == "parasite" ~ "Basidiomycota",
  raw_resourceRelation == "host"     ~ ""))
```

### order

```{r}
taxon %<>% mutate(order = case_when(
  raw_resourceRelation == "parasite" ~ "Uredinales",
  raw_resourceRelation == "host"     ~ ""))
```

### family

```{r}
taxon %<>% mutate(family = raw_family) %<>% replace_na(replace = list(family = "")) #' Remove NA's for host species
```

### genus

```{r}
taxon %<>% mutate(genus = pn_genusorabove) %<>% replace_na(replace = list(genus = "")) #' Remove NA's for host species
```

### specificEpithet

```{r}
taxon %<>% mutate(specificEpithet = pn_specificepithet) %<>% replace_na(replace = list(specificEpithet = "")) #' Remove NA's for host species
```

### infraspecificEpithet

```{r}
taxon %<>% mutate(infraspecificEpithet = pn_infraspecificepithet) %<>% replace_na(replace = list(infraspecificEpithet = "")) #' Remove NA's for host species
```

### taxonRank

Information for `taxonRank` is contained in `pn_rankmarker`:

```{r}
taxon %>% distinct(pn_rankmarker)
```

Generate `taxonRank`by recoding abbreviations:

```{r}
taxon %<>% mutate(taxonRank = recode(pn_rankmarker,
  "sp." = "species",
  "var." = "variety",
  "subsp." = "subspecies",
  "fam." = "family",
  "f." = "form",
  .missing = ""))
```

### scientificNameAuthorship

The format specifications for `scientificNameAuthorship` is given [here](http://rs.tdwg.org/dwc/terms/history/index.htm#scientificNameAuthorship-2009-04-24).The mapping is as follows:
* no (bracket)authorship provided: `scientificNameAuthorship` is left blank.
* only authorship is provided, no information on bracketauthorship: we copy values in `pn_authorship`..
* both authorship and bracketauthorship is provided: the required format here is (bracketauthorship) authorship.

```{r}
taxon %<>% mutate(scientificNameAuthorship = case_when(
  pn_bracketauthorship != "" & pn_authorship != "" ~ paste( paste0("(", pn_bracketauthorship, ")"), pn_authorship, sep =" "),
  pn_bracketauthorship == "" & pn_authorship != "" ~ pn_authorship,
  TRUE ~ ""))
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(nomenclaturalCode = "ICBN")
```

## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("raw_"), -starts_with("pn_"))
```

Preview data:

```{r}
head(taxon)
```

Save to CSV:

```{r}
write.csv(taxon, file = dwc_taxon_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create Distribution extension

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

## Pre-processing

The distribution applies to rust fungi only. We remove the host plants from `raw_data` and create `distribution`:

```{r}
distribution <- raw_data %>% filter(raw_resourceRelation == "parasite")
```

## Term mapping

### taxonID

```{r}
distribution %<>% mutate(taxonID = raw_taxon_id)
```

### locationID

```{r}
distribution %<>% mutate(locationID = "ISO_3166-2:BE")
```

### locality

```{r}
distribution %<>% mutate(locality = "Belgium")
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = "BE")
```

### occurrenceStatus

Information for `occurrenceStatus` is contained in `raw_occurrenceStatus`:

```{r}
distribution %>% distinct(raw_occurrenceStatus)
```

These values are conform the [IUCN definitions](http://www.iucnredlist.org/technical-documents/red-list-training/iucnspatialresources).

```{r}
distribution %<>% mutate(occurrenceStatus = raw_occurrenceStatus)
```

### establishmentMeans

establishmentMeans information is contained in `raw_Column`:

```{r}
distribution %>% distinct(raw_Column)
```

`native` is not part of the [establishmentMeans vocabulary](http://rs.gbif.org/vocabulary/gbif/establishment_means.xml). Here, we use `introduced`:

```{r}
distribution %<>% mutate(establishmentMeans = recode(raw_Column,
  "native" = "introduced",
  .missing = ""))
```

### eventDate

eventDate information is contained in `raw_from` (date of first observation) and `raw_to` (date of last observation). We will map this information as a ISO 8601 date range (start_date / end_date). For this, we need to clean and merge the information in `raw_from` and `raw_to`.

`raw_from` contains the information for the first observation. At this point, the following date formats are given: `dd.m.yyyy`, `m.yyyy`, `m/yyyy` and `yyyy`, with the information for the days and years expressed as arabic numericals and for the months as arabic or roman numericals. We need to split `raw_from` to convert the roman numericals.

```{r}
distribution %<>% separate(raw_from, 
                           into = c("day", "month", "year"),
                           remove = FALSE,
                           extra = "merge",
                           fill = "left")     
```

Convert roman to arabic numericals in  `month`:

```{r}
distribution %<>% mutate(month = recode(month,
    "I"   = "01",
    "II"  = "02",
    "III" = "03",
    "IV"  = "04",
    "V"   = "05",
    "VI"  = "06",
    "VII" = "07",
    "VIII"= "08",
    "IX"  = "09",
    "X"   = "10",
    "XI"  = "11",
    "XII" = "12"))
```

We remerge the year, month and day information for the date of first observation and express this in a ISO 8601 format. we need to add the leading zeros to months (using `sprintf`), we combine year, month and date information with "-" (using `paste`) and we strip the trailing "-NA" if day or month is "NA" (using `gsub`):

```{r}
distribution %<>% mutate(start_date = 
  gsub("-NA", "",
       paste(year, 
             sprintf("%02d", as.integer(month)),
             sprintf("%02d", as.integer(day)),
             sep = "-")))
```

Compare the first records of `start_date` with `raw_from`:

```{r}
distribution %>% select(raw_from, start_date) %>% head(n = 20)
```

remove `day`, `month` and `year`:

```{r}
distribution %<>% select(-day, -month, -year)
```


The information for the date of the last observation is contained in `raw_to`.
The same principles and steps can be repeated for the mapping of `raw_to` into `end_date`

Separate `raw_to` into `day`, `month` and `year`:

```{r}
distribution %<>% separate(raw_to, 
                           into = c("day", "month", "year"),
                           remove = FALSE,
                           extra = "merge",
                           fill = "left")     
```

Convert roman to arabic numericals in  `month`:

```{r}
distribution %<>% mutate(month = recode(month,
    "I"   = "01",
    "II"  = "02",
    "III" = "03",
    "IV"  = "04",
    "V"   = "05",
    "VI"  = "06",
    "VII" = "07",
    "VIII"= "08",
    "IX"  = "09",
    "X"   = "10",
    "XI"  = "11",
    "XII" = "12"))
```

we need  add the leading zeros to months (using `sprintf`), we combine year, month and date information with "-" (using `paste`) and we strip the trailing "-NA" if day or month is "NA" (using `gsub`):

```{r}
distribution %<>% mutate(end_date = 
  gsub("-NA", "",
       paste(year, 
             sprintf("%02d", as.integer(month)),
             sprintf("%02d", as.integer(day)),
             sep = "-")))
```

Compare the first records of `end_date` with `raw_from`:

```{r}
distribution %>% select(raw_to, end_date) %>% head(n = 20)
```

remove `day`, `month` and `year`:

```{r}
distribution %<>% select(-day, -month, -year)
```

Merge `start_date` and `end_date` in `eventDate`:

```{r}
distribution %<>% mutate(eventDate = paste(start_date, end_date, sep = "/"))
```

Remove "/NA" or "NA/NA":

```{r}
distribution %<>% mutate(eventDate = gsub("([/][N][A]|([NA]{2}[/][NA]{2}))", "", eventDate))
```

### source

```{r}
distribution %<>% mutate(source = raw_bibliographicCitation)
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("raw_"), -start_date, -end_date)
```

Preview data:

```{r}
head(distribution)
```

Save to CSV:

```{r}
write.csv(distribution, file = dwc_distribution_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create Resource Relationship extension

Map the data to [Resource Relationship](http://rs.gbif.org/extension/dwc/resource_relation_2018_01_18.xml).

```{r}
resource_relationship <- raw_data
```

In this extension, we map all interactions between the rust fungi and host plants, with one line representing a one-to-one taxon interaction. These interactions are saved in `interactions`. We add the one-to-one taxon interactions to `resource_relationship`: 

```{r}
resource_relationship %<>% right_join(interactions, by = c("raw_scientificName" = "scientificName_parasite"))
```

### resourceID

This Darwin Core term refers to the taxonID of the _object_ of the relationship. This is the taxonID of the host plants. By joining `resource_relationship` with `interactions` on the the scientificName of the parasite, the taxonID's of the host plants were removed from the `resource_relationship`. We need to extract them from `raw_data`:

```{r}
taxonID_host <- raw_data %>% filter(raw_resourceRelation == "host") %>% 
  select(raw_scientificName, raw_taxon_id)
```

Join `resource_relationship` with `taxonID_host`:

```{r}
resource_relationship %<>% left_join(taxonID_host, by = c("scientificName_host" = "raw_scientificName"))
```

Rename `raw_taxon_id.y`:

```{r}
resource_relationship %<>% rename("resourceID" = "raw_taxon_id.y")
```

### relatedResourceID

This Darwin Core term refers to the taxonID of the _subject_ of the relationship. This is the taxonID of the rust fungi.

```{r}

```

### relationshipOfResource	

```{r}
resource_relationship %<>% mutate(relationshipOfResource = "parasite_of")
```

### relationshipAccordingTo	

```{r}
resource_relationship %<>% mutate(relationshipAccordingTo = raw_bibliographicCitation)
```

## Post-processing:

Remove the original columns:

```{r}
resource_relationship %<>% select(-scientificName_parasite, -scientificName_host, -raw_bibliographicCitation)
```

Preview data:

```{r}
head(resource_relationship)
```

Save to CSV:

```{r}
write.csv(resource_relationship, file = dwc_relationship_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```



