---
title: "Darwin Core mapping"
subtitle: "For dataset: Checklist of the rust fungi of Belgium"
author: 
- Lien Reyserhove
- Peter Desmet
- Quentin Groom
- Dimitri Brosens
date:"`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

The data in the raw file has already been mapped to Darwin Core terms. 
This document describes how we further process the data to render the Taxon Core file and the different Extensions.

# Setup

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r, eval = F}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # For data transformations

# None core tidyverse packages:
library(magrittr)  # For %<>% pipes

# Other packages
library(janitor)   # For cleaning input data
library(digest)    # Create hash function digests
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/checklist of rusts.tsv"

# Processed files:

```

# Read data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read.table(raw_data_file, header = T, sep = "\t") 

# Clean the data somewhat: remove empty rows if present
raw_data %<>%
  remove_empty_rows()     # Remove empty rows
```

